---
title: "README"
author: "npc"
output: html_document
---

# Questions

Your data analysis must address the following questions:

## Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

## Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

The analysis document must have at least one figure containing a plot, no more than three figures.

# Layout

## Title
briefly summarizes your data analysis

## Synopsis
describes and summarizes your analysis in at most 10 complete sentences.

## Data Processing
how the data were loaded into R and processed for analysis.

### Read storm data from csv file
```{r, echo=TRUE, cache = TRUE}
source("storm_data.R")
data = read_raw();
```

### cleanup data
- all lower case
- replace sign with single space
```{r, echo=TRUE, cache = TRUE}
data = cleanup(data)
```

### Definition of data
*Question1:* Population health damage by event is calculated from 
```{r, echo=TRUE, cache = TRUE}
population_damage = as.formula("FATALITIES + INJURIES ~ EVTYPE")
```

*Question2:* Economic damage by event is calculated from 
```{r, echo=TRUE, cache = TRUE}
economic_damage = as.formula("PROPDMG + CROPDMG ~ EVTYPE")
```

## Results

### Calculate pulation health damage by event
```{r, echo=TRUE, cache = TRUE}
agg_population = aggregate(population_damage, data = data, FUN = sum)
agg_population = filter_result(agg_population)
```

### Calculate economic damage by event
```{r, echo=TRUE, cache = TRUE}
agg_economic = aggregate(economic_damage, data = data, FUN = sum)
agg_economic = filter_result(agg_economic)
```

### Plot top 10 of harmful events to population health
bla bla bla
```{r, echo=TRUE, cache = TRUE}
library(ggplot2)
ggplot(data=agg_population, 
    aes(x=EVTYPE, y=agg_population[[2]], fill=EVTYPE)) +
    geom_bar(colour="black", stat="identity") +
    guides(fill=FALSE) +
    xlab("Event type") +
    ylab("Number of fatalities and injuries")
```

### Plot top 10 of harmful events to economic
bla bla bla
```{r, echo=TRUE, cache = TRUE}
ggplot(data=agg_economic, 
    aes(x=EVTYPE, y=agg_economic[[2]] / 10^9, fill=EVTYPE)) +
    geom_bar(colour="black", stat="identity") +
    guides(fill=FALSE) +
    xlab("Event type") +
    ylab("Property and crop damage in billion $")
```





